---
title: "Effects of COVID-19 on the Academic Performance of College Students"
subtitle: "Figures"
date: today
format:
  docx:
    reference-doc: tf_template.docx
---


```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE, message=FALSE, warning=FALSE}

if (!require("pacman")) {
  install.packages("pacman", repos = "http://cran.us.r-project.org")
}

# Load and install all packages at once
pacman::p_load(
  tidymodels,
  tidyverse,
  rio,
  naniar,
  labelled,
  sjlabelled,
  haven,
  fixest,
  vtable,
  ggfixest,
  modelsummary,
  gtsummary,
  simputation,
  kableExtra,
  readxl,
  panelsummary,
  cowplot,
  parameters,
  patchwork
)

# Set a single global theme, font family and base size for all plots
theme_set(theme_minimal(base_size = 12, base_family = "serif"))

# Reusable plot theme (use for explicit overrides)
plot_theme <- theme_minimal(base_size = 12, base_family = "serif") +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(size = 12, face = "plain"),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10)
  )

# Colorblind-friendly palette (Okabe–Ito) with lighter tints for labels/secondary bars
# Primary hues (high contrast) and lighter variants for two-level comparisons
primary_fill  <- "#0072B2"  # blue (primary)
primary_light <- "#7FB3D6"  # lighter blue (tint)
primary_dark  <- "#004F80"  # darker blue

# Secondary contrasting hues (use for stacked bars / multiple groups)
secondary1      <- "#009E73"  # green
secondary1_light<- "#8FD8BF"
secondary1_dark <- "#006B4E"

secondary2      <- "#D55E00"  # vermillion / orange
secondary2_light<- "#FFB58A"
secondary2_dark <- "#8A3600"

secondary3      <- "#CC79A7"  # pink / magenta (optional)
secondary3_light<- "#E7A5C8"
secondary3_dark <- "#8A3A61"

# Named palettes for stacked/filled measures (keep semantics consistent across plots)
withdrawal_palette <- c(
  "avg_NC" = secondary1,       # No Credit -> green
  "avg_CR" = secondary2,       # Credit    -> orange
  "avg_WD" = primary_fill      # Withdrawal-> blue (primary)
)

gpa_type_palette <- c(
  "High GPA" = primary_light,  # lighter primary for 'High'
  "Low GPA"  = primary_dark    # darker primary for 'Low'
)

inst_mode_palette <- c(
  "Hybrid" = secondary1_light, # lighter green for Hybrid
  "Online" = secondary2_dark   # darker orange for Online
)

# Shared neutral color for text, errorbars, vlines and outlines
plot_neutral <- "#333333"
```

\pagebreak

```{r data-load, echo = FALSE, warning=FALSE, message=FALSE, include = FALSE}
### Loading the question level data ------------------------------------------

question_url <- "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/question_level.rds"
question_level <- read_rds(question_url)

### Loading the exam level data ------------------------------------------
## These scores are out of possible 40 points.

exam_url <- "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/exam_level.rds"
exam_level <- read_rds(exam_url)


### Loading the course withdrawal data ------------------------------------------
## It has share of students who took withdrawal, CR, NC options.
full_grades <- read_dta(
  "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/grades-by-sem.dta"
) # s2020 available
full_g <- full_grades |>
  mutate(
    year = if_else(
      semester == "S19" | semester == "F19",
      2019,
      if_else(
        semester == "S20" | semester == "F20",
        2020,
        if_else(semester == "S21" | semester == "F21", 2021, 2022)
      )
    )
  ) |>
  mutate(
    session = factor(if_else(
      semester %in% c("S19", "S20", "S21", "S22"),
      "S",
      "F"
    ))
  ) |>
  mutate(session = fct_relevel(session, c("S", "F")))

```

::: {custom-style="Figure caption"}
Figure 1: Average final GPA in ECO 1001 across semesters
:::

```{r gpa-across-semesters-eco-1001, echo = FALSE, warning=FALSE, message=FALSE, fig.pos="H",  fig.align = "center"}
# Define semester labels
gpa_labels <- c(
  "Spring 2019",
  "Fall 2019",
  "Spring 2020",
  "Fall 2020",
  "Spring 2021",
  "Fall 2021",
  "Spring 2022"
)

# Summarize GPA statistics by semester
gpa_summary <- full_g |>
  dplyr::select(year, session, coursegpa) |>
  dplyr::group_by(year, session) |>
  dplyr::summarise(
    avg_gpa = mean(coursegpa, na.rm = TRUE),
    sd_gpa = sd(coursegpa, na.rm = TRUE),
    n = dplyr::n(),
    .groups = 'drop'
  ) |>
  dplyr::mutate(
    margin_error = 1.96 * (sd_gpa / sqrt(n)),
    semester = factor(seq_along(year), labels = gpa_labels)
  )

# Plot the results
ggplot(gpa_summary, aes(x = semester, y = avg_gpa, label = round(avg_gpa, 2))) +
  geom_bar(stat = "identity", width = 0.3, fill = primary_fill, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = avg_gpa - margin_error, ymax = avg_gpa + margin_error),
    width = 0.1,
    color = "darkblue"
  ) +
  geom_text(
    size = 3,
    vjust = -2.5,
    color = "black",
    family = "serif"
  ) +
  ylim(0, 4) +
  labs(
    x = NULL,
    y = "Course GPA",
    title = ""
  ) +
  scale_x_discrete(
    labels = gpa_labels,
    guide = ggplot2::guide_axis(n.dodge = 2)
  ) +
  plot_theme
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 2: Withdrawal rates in ECO 1001 across semesters
:::

```{r withdrawal-rate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos="H", fig.align = "center"}
## Figure 2: Withdrawal rates in ECO 1001 across semesters ------------------

## 376 students withdrew + 39 students withdrew
## 296 students CR
## 55 students NC
## 160 missing coursegpa

# Semester labels
withdrawal_labels <- c(
  "Spring 2019",
  "Fall 2019",
  "Spring 2020",
  "Fall 2020",
  "Spring 2021",
  "Fall 2021",
  "Spring 2022"
)

# Summarize withdrawal-related measures
withdrawal_summary <- full_g |>
  dplyr::select(year, session, CR, NC, wd) |>
  dplyr::group_by(year, session) |>
  dplyr::summarise(
    avg_NC = mean(NC, na.rm = TRUE),
    avg_CR = mean(CR, na.rm = TRUE),
    avg_WD = mean(wd, na.rm = TRUE),
    .groups = 'drop'
  ) |>
  dplyr::mutate(
    semester = factor(seq_along(year), labels = withdrawal_labels)
  ) |>
  tidyr::pivot_longer(
    cols = c("avg_NC", "avg_CR", "avg_WD"),
    names_to = "measure",
    values_to = "avg"
  ) |>
  dplyr::mutate(avg = 100 * avg)

# Set up color scheme for measures
withdrawal_colors <- c(
  "avg_NC" = "#0072B2", # blue
  "avg_CR" = "#D55E00", # vermillion / orange
  "avg_WD" = "#009E73" # green (distinct from blue & orange)
)


# Plot the withdrawal rates
ggplot(
  withdrawal_summary,
  aes(x = semester, y = avg, fill = measure, label = paste0(round(avg, 2), "%"))
) +
  geom_col(width = 0.5, position = position_stack(), alpha = 0.7) +
  geom_text(
    size = 2,
    vjust = -2.5,
    position = position_stack(0.),
    color = "black",
    family = "serif"
  ) +
  ylim(0, 50) +
  scale_fill_manual(
    values = withdrawal_palette,
    labels = c("No Credit", "Credit", "Withdrawal")
  ) +
  labs(
    x = NULL,
    y = "Withdrawal, Credit, No Credit Rate (%)",
    title = ""
  ) +
  scale_x_discrete(
    labels = withdrawal_labels,
    guide = ggplot2::guide_axis(n.dodge = 2)
  ) +
  plot_theme
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 3: Dynamic effects of COVID-19 on students' performance
:::

```{r dynamic-effects, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=5, dpi=150, fig.retina=1, out.width="70%", fig.align="center"}
 
#####################################################################################
############### LONG TERM EFFECTS ON HIGH VS LOW GPA USING YEAR DUMMIES #############
#####################################################################################

exam_longterm_gpa <- exam_level |>
  feols(fml = score ~ i(factor_var = time, var = lowgpa, ref = 1) + online + female + r_hispa + r_black + r_asian + r_other
    + session + parttime + gpamiss | time + instructor, vcov = "hc1")

ques_longterm_gpa <- question_level |>
  feols(fml = correct ~ i(factor_var = time, var = lowgpa, ref = 1) + online + female + r_hispa + r_black + r_asian + r_other
    + session + parttime + gpamiss | time + instructor, vcov = "hc1")

#####################################################################################
############### LONG TERM EFFECTS ON ONLINE VS HYBRID USING YEAR DUMMIES ###########
#####################################################################################

exam_longterm_online <- exam_level |>
  filter(time %in% c("1", "2", "3")) |>
  feols(fml = score ~ i(factor_var = time, var = online, ref = 1) + cumgpa + female + r_hispa + r_black + r_asian + r_other
    + parttime + gpamiss | instructor, vcov = "hc1")

ques_longterm_online <- question_level |>
  filter(time %in% c("1", "2", "3")) |>
  feols(fml = correct ~ i(factor_var = time, var = online, ref = 1) + cumgpa + female + r_hispa + r_black + r_asian + r_other
    + parttime + gpamiss | instructor, vcov = "hc1")

# Create plots with consistent theme, fonts and colors
plot_gpa_exam <- ggiplot(exam_longterm_gpa, 
                         main = "Exam Scores: Low vs High GPA",
                         xlab = "Time Period",
                         ylab = "Estimates",
                         ref.line = FALSE,
                         pt.join = FALSE,
                         geom_style = "pointrange",
                         ci_level = 0.95) +
  geom_vline(xintercept = 1, linetype = "dashed", color = plot_neutral) +
  plot_theme +
  theme(plot.title = element_text(size = 10, face = "plain"),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 9, family = "serif"))

plot_gpa_ques <- ggiplot(ques_longterm_gpa, 
                         main = "Matched Questions: Low vs High GPA",
                         xlab = "Time Period",
                         ylab = "Estimates",
                         ref.line = FALSE,
                         pt.join = FALSE,
                         geom_style = "pointrange",
                         ci_level = 0.95) +
  geom_vline(xintercept = 1, linetype = "dashed", color = plot_neutral) +
  plot_theme +
  theme(plot.title = element_text(size = 10, face = "plain"),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 9, family = "serif"))

plot_online_exam <- ggiplot(exam_longterm_online, 
                            main = "Exam Scores: Online vs Hybrid",
                            xlab = "Time Period",
                            ylab = "Estimates",
                            ref.line = FALSE,
                            pt.join = FALSE,
                            geom_style = "pointrange",
                            ci_level = 0.95) +
  geom_vline(xintercept = 1, linetype = "dashed", color = plot_neutral) +
  plot_theme +
  theme(plot.title = element_text(size = 10, face = "plain"),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 9, family = "serif"))

plot_online_ques <- ggiplot(ques_longterm_online, 
                            main = "Matched Questions: Online vs Hybrid",
                            xlab = "Time Period",
                            ylab = "Estimates",
                            ref.line = FALSE,
                            pt.join = FALSE,
                            geom_style = "pointrange",
                            ci_level = 0.95) +
  geom_vline(xintercept = 1, linetype = "dashed", color = plot_neutral) +
  plot_theme +
  theme(plot.title = element_text(size = 10, face = "plain"),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 9, family = "serif"))

(plot_gpa_exam + plot_gpa_ques) / 
  (plot_online_exam + plot_online_ques)
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 1A: Average final exam scores in ECO 1001 across semesters
:::

```{r exam-scores-mean, echo=FALSE, warning=FALSE, message=FALSE, fig.pos="H", fig.width = 6, fig.height = 4, fig.align = "center"}
### Figure A1: Average final exam scores in ECO 1001 across semesters ------------------

# Semester labels for exam scores

exam_labels <- c(
  "Spring 2019", "Fall 2019", "Fall 2020",
  "Spring 2021", "Fall 2021", "Spring 2022"
)

# Summarize exam score statistics

exam_summary <- exam_level|>
  dplyr::select(score, year, session)|>
  dplyr::mutate(session = forcats::fct_relevel(session, c("S", "F")))|>
  dplyr::group_by(year, session)|>
  dplyr::summarise(
    avg_score = mean(score * 0.4, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    n = dplyr::n(),
    std_error = sd_score / sqrt(n),
    .groups = 'drop'
  )|>
  dplyr::mutate(
    semester = factor(seq_along(year), labels = exam_labels),
    margin_error = 1.96 * std_error
  )

# Plot exam scores across semesters

ggplot(exam_summary, aes(x = semester, y = avg_score, label = round(avg_score, 2))) +
  geom_bar(stat = "identity", width = 0.3, fill = primary_fill, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = avg_score - margin_error, ymax = avg_score + margin_error),
    width = 0.2, color = "darkblue"
  ) +
  geom_text(
    size = 3, vjust = -3.5, color = "black", family = "serif"
  ) +
  ylim(0, 50) +
  labs(
    x = NULL,
    y = "Average Final Exam Score",
    title = ""
  ) +
  scale_x_discrete(labels = exam_labels,
  guide = guide_axis(n.dodge = 2)) +
  plot_theme
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 2A: Share of High vs Low GPA Students
:::

```{r share-gpa-low-high, echo=FALSE, warning=FALSE, message=FALSE, fig.pos="H", fig.width = 6, fig.height = 4, fig.align = "center"}
### Figure A2: Student shares in low and high GPA groups ------------------
## S2019,F2019,F2020,S2021,F2021,S2022
## S2019/F2019 combined due to small sample size

# Semester labels for GPA share plot
gpa_share_labels <- c(
  "Spring/Fall 2019", "Fall 2020",
  "Spring 2021", "Fall 2021", "Spring 2022"
)


# Prepare summary data: student share by GPA type
stu_share_gpa_summary <- question_level|>
  dplyr::distinct(id, .keep_all = TRUE)|>
  dplyr::group_by(time, id, typegpa)|>
  dplyr::count()|>
  dplyr::group_by(time, typegpa)|>
  dplyr::count(name = "n")|>
  dplyr::group_by(time)|>
  dplyr::mutate(share = n / sum(n))|>
  dplyr::ungroup()

# Color palette for GPA types
gpa_type_colors <- c("High GPA" = "skyblue", "Low GPA" = "darkblue")

stu_share_gpa_summary <- stu_share_gpa_summary |>
  dplyr::mutate(typegpa = dplyr::recode(typegpa,
                                      "high" = "High GPA",
                                      "low"  = "Low GPA"))

# Plot share of students by GPA type
ggplot(stu_share_gpa_summary, aes(x = factor(time), y = share, fill = typegpa, label = round(share, 2))) +
  geom_col(
    width = 0.5,
    alpha = 0.8,
    position = position_dodge(0.6),
    color = "black"
  ) +
  geom_text(
    size = 3,
    vjust = -1,
    position = position_dodge(0.6),
    family = "serif"
  ) +
  ylim(0, 1) +
  labs(
    x = NULL,
    y = "Share of High and Low GPA Students",
    fill = "GPA Type",
    title = ""
  ) +
  scale_fill_manual(values = gpa_type_palette) +
  scale_x_discrete(labels = gpa_share_labels,
  guide = guide_axis(n.dodge = 2)) +
  plot_theme
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 3A: Average GPA in High vs Low GPA Group of Students
:::

```{r meangpa-gpa-group, echo=FALSE, warning=FALSE, message=FALSE, fig.pos="H", fig.width = 6, fig.height = 3.5, fig.align = "center"}
### Figure A3: Average GPA in low and high GPA groups in ECO 1001 across semesters ------------------
## S2019,F2019,F2020,S2021,F2021,S2022
# Prepare semester labels
mean_gpa_labels <- c(
  "Spring/Fall 2019", "Fall 2020",
  "Spring 2021", "Fall 2021", "Spring 2022"
)

# Summarize mean cumulative GPA by group
stu_mean_gpa_summary <- question_level %>%
  dplyr::distinct(id, .keep_all = TRUE) %>%
  dplyr::select(time, cumgpa, typegpa) %>%
  dplyr::group_by(time, typegpa) %>%
  dplyr::summarise(
    avg_gpa = mean(cumgpa, na.rm = TRUE),
    sd_score = sd(cumgpa, na.rm = TRUE),
    n = dplyr::n(),
    std_error = sd_score / sqrt(n),
    .groups = "drop"
  ) |>
  dplyr::mutate(margin_error = 1.96 * std_error)

# GPA type colors (ensure levels match your data)
gpa_type_colors <- c("High GPA" = "skyblue", "Low GPA" = "darkblue")

# If your data uses values like 'high'/'low', relabel before plotting
stu_mean_gpa_summary <- stu_mean_gpa_summary %>%
  dplyr::mutate(typegpa = dplyr::recode(typegpa,
                                        "high" = "High GPA",
                                        "low"  = "Low GPA"))

# Plot mean GPA by group across semesters
ggplot(stu_mean_gpa_summary, aes(
  x = factor(time),
  y = avg_gpa,
  group = typegpa,
  fill = typegpa,
  label = round(avg_gpa, 2)
)) +
  geom_col(
    width = 0.5,
    alpha = 0.8,
    position = position_dodge(0.6),
    color = "black"
  ) +
  geom_text(
    size = 3,
    vjust = -1,
    position = position_dodge(0.75),
    color = "black",
    family = "serif"
  ) +
  ylim(0, 6) +
  labs(
    x = NULL,
    y = "Mean GPA of High and Low GPA Students",
    fill = "GPA Type",
    title = ""
  ) +
  scale_x_discrete(
    breaks = c("1", "2", "3", "4", "5"),
    labels = mean_gpa_labels,
    guide = guide_axis(n.dodge = 2)
  ) +
  scale_fill_manual(values = gpa_type_palette) +
  plot_theme
```

\pagebreak

::: {custom-style="Figure caption"}
Figure 4A: Share of the Students in Hybrid vs Online Classes
:::

```{r student-share-online-hybrid, echo=FALSE, warning=FALSE, message=FALSE,fig.pos="H", fig.width = 6, fig.height = 5, fig.align = "center"}
### Figure 4A: Student shares in hybrid and online classes ------------------
#### S2019,F2019,F2020,S2021,F2021,S2022

# Semester labels for instruction mode plot
inst_mode_labels <- c(
  "Spring/Fall 2019", "Fall 2020",
  "Spring 2021", "Fall 2021", "Spring 2022"
)

# Prepare summary data: student share by instruction mode
stu_share_inst_summary <- question_level %>%
  dplyr::distinct(id, .keep_all = TRUE) %>%
  dplyr::group_by(time, id, instruction_mode) %>%
  dplyr::count() %>%
  dplyr::group_by(time, instruction_mode) %>%
  dplyr::count(name = "n") %>%
  dplyr::group_by(time) %>%
  dplyr::mutate(share = n / sum(n)) %>%
  dplyr::ungroup()

# Set color palette for instruction modes -- ensure it matches your data
inst_mode_colors <- c("Hybrid" = "skyblue", "Online" = "darkblue")

# Relabel for legend, if your values are different
stu_share_inst_summary <- stu_share_inst_summary %>%
  dplyr::mutate(instruction_mode = dplyr::recode(instruction_mode,
                                                 "H" = "Hybrid",
                                                 "O" = "Online"
  ))

# Plot student share by instruction mode
ggplot(stu_share_inst_summary, aes(
  x = factor(time),
  y = share,
  fill = instruction_mode,
  label = round(share, 2)
)) +
  geom_col(
    width = 0.2,
    alpha = 0.7,
    position = position_dodge(0.4),
    color = "black"
  ) +
  geom_text(
    size = 3,
    vjust = -1,
    position = position_dodge(0.75),
    family = "serif"
  ) +
  ylim(0, 1) +
  labs(
    x = NULL,
    y = "Share of Students in Hybrid and Online Classes",
    fill = "Instruction Mode",
    title = ""
  ) +
  scale_x_discrete(
  breaks = c("1", "2", "3", "4", "5"),
  labels = inst_mode_labels,
  guide = guide_axis(n.dodge = 2)) +
  scale_fill_manual(values = inst_mode_palette) +
  plot_theme
```

