---
title: "Effects of COVID-19 on the Academic Performance of College Students"
author:
  - name: Aman Desai
    affiliation: Quinnipiac University
    email: aman.desai@quinnipiac.edu
date: today
format:
  docx:
    reference-doc: tf_template.docx  
bibliography: references.bib
biblio-style: natbib
abstract: |
  I analyze the impact of the COVID-19 pandemic on undergraduates’ performance in an introductory economics course at a large public university. One challenge in analyzing student academic outcomes during the pandemic was the explicit change in grading policies by college administrators as well as the implicit adjustment by faculty designed to mitigate the impact of an abrupt shift to online learning amidst the stress and uncertainly associated with the pandemic. To limit the impact of grading policies I analyze changes in the raw scores on a common final administered to all sections of the course the year before and for four semesters after the spring of 2020. To limit variation in the difficulty of the exams from before to during the pandemic, I compare student performance on nearly identical questions on the final exam overtime. Adjusted mean scores on the common final fell by a point and the probability of answering the qualitatively same question on the final fell, on average, by 1.5 percentage points. Students with lower GPAs were 3.3 percentage points (or 0.02 standard deviations) less likely to answer similar questions correctly relative to students with higher GPAs during the pandemic. Also, the mean probability of answering a nearly identical question before and after suddenly moving to online classes increased by 5.6 percentage points.
  
  **Keywords:** COVID-19, academic performance, undergraduate education, online learning
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE, message=FALSE, warning=FALSE}

if (!require("pacman")) {
  install.packages("pacman", repos = "http://cran.us.r-project.org")
}

# Load and install all packages at once
pacman::p_load(
  tidymodels,
  tidyverse,
  rio,
  naniar,
  labelled,
  sjlabelled,
  haven,
  fixest,
  vtable,
  ggfixest,
  modelsummary,
  gtsummary,
  simputation,
  kableExtra,
  readxl,
  panelsummary,
  cowplot,
  parameters,
  patchwork
)

# Set default theme
theme_set(theme_bw(base_family = "serif"))

```

\pagebreak

# Introduction

The COVID-19 pandemic of March 2020 was disruptive across many domains, with higher education being one of them. Policies were implemented worldwide in response to this global crisis, resulting in changes in the educational setting. Educational instructions were abruptly moved online without prior preparation. This had a negative effect on primary and secondary education, leading to significant learning loss for students [@grewenig_covid-19_2021; @fuchs-schundeln_covid-induced_2022].

Although the socioeconomic consequences of COVID-19 have been extensively studied from various perspectives, research on the impact of the pandemic on college students remains limited and yields conflicting results. Most studies examining the impact of the pandemic on students’ academic performance measure outcomes such as GPA and course completion rates. Although useful, these measures are confounded by the numerous responses of students, faculty, and administrators during the pandemic. For example, cheating became a challenging issue in the rapid move to online teaching [@ives_did_2024; @jenkins_when_2023;  @walsh_why_2021]. The faculty adopted more lenient grading practices and reduced exam difficulties. Administrators altered grading policies regarding course withdrawals and pass/fail options [@rodriguez-planas_covid-19_2022]. These responses made comparisons with pre-pandemic test scores, or the term GPA less reliable for quantifying learning loss during the pandemic in college students.

I overcome these assessment and grading issues using unique exam-level data from a large public university in New York City. First, I analyze students' performance on final exams before and during the pandemic in an *introductory microeconomics* course. Approximately 800 students across ten sections of the course attempted a common final exam each semester. This reduces the variation in the difficulty of exams across sections. However, the difficulty of an exam may have changed in response to the pandemic. Thus, I compare students’ performance on specific exam questions that were qualitatively almost identical before and during the pandemic by matching questions from answer sheets from the common final exams before and during the pandemic. By focusing on students’ performance on nearly identical questions before and during the pandemic, I remove the variation in outcomes due to possible changes in the difficulty of these exams during the pandemic. By combining the matched question-level data with student characteristics, I estimated how the pandemic affected students’ average probability of correctly answering similar questions from pre-pandemic common exams during the crisis.

I begin with a before and after analysis, adjusting for student characteristics, time periods, and instructor fixed effects. I argue that more capable students are more likely to adjust to online instruction more effectively. Thus, I use a difference-in-difference design and compared students with pre-course GPA above (high GPA) and below (low GPA) the median before and during the pandemic. I observe that during the pandemic, low-GPA students were less likely to answer qualitatively similar questions from the pre-pandemic exams relative to students with higher GPAs. My analysis of dynamic effects reveals that by spring 2022, the performance gap persisted between low and high GPA students, both in overall exam scores and in their likelihood of correctly answering nearly identical questions compared with pre-pandemic levels.

I also analyze the students' performance on matched questions by difficulty level. I find no statistically significant impact of the pandemic on low GPA students' average probability of answering nearly identical "easy" questions correctly, but I find significant effect on their performance with nearly identical *hard* questions. I also provide a similar analysis comparing outcomes between students enrolled in online and hybrid classes.
My findings align with existing research on learning loss during this period. By analyzing the performance of qualitatively similar exam questions before and during the pandemic, I contribute to the literature by offering more reliable estimates of learning loss compared to traditional metrics, such as GPA and course withdrawals.

The next section reviews the current literature on the effects of the pandemic on college students’ academic outcomes. Section 3 discusses the data, section 4 explains the estimation strategy, section 5 reports the results, and section 6 concludes the paper.

# Literature Review

Most early studies analyzing the impact of COVID-19 on undergraduate student outcomes were based on surveys about their experiences during the pandemic. @jaeger_global_2021 was the first to document the negative impact of the COVID-19 pandemic using surveys administered to university students in 28 universities in the United States, Spain, Australia, Sweden, Austria, Italy, and Mexico between April and October 2020. Their preliminary results reported disparate impacts on different socio-economic and demographic groups. @aucejo_impact_2020, one of the first papers studying the effect of COVID-19 on college student outcomes, surveyed 1,500 students at a large public university in the United States. They found significant negative effects of the pandemic on student outcomes. Due to the pandemic, 13% of students delayed graduation, 40% lost a job, internship, or offer, and 29% expected an earnings loss by age 35. They also found large disparate impacts of the pandemic across socio-economic statuses. Lower-income students were 55% more likely than their higher-income peers to have delayed graduation due to COVID-19. 

Along the same lines, @rodriguez-planas_hitting_2020 collected data on students’ experiences during the pandemic using an online survey at an urban public college in New York City in the summer of 2020. The author found significant disruptions in students’ lives due to the pandemic. Because of COVID, between 14% and 34% of students considered dropping a class during spring 2020, 30% modified their graduation plans, and the freshman fall retention rate dropped by 26%. The pandemic also deprived 39% of students of their jobs, while 35% of students saw their earnings reduced. Pell grant recipients (students from lower-income families) were 20% more likely to lose a job due to the pandemic and 17% more likely to experience earning losses than non-Pell recipients. Other vulnerable groups, such as first-generation and transfer students, were relatively more affected. Since they seem to rely less on financial aid and more on income from wage and salary jobs, both their educational and employment outcomes were more negatively impacted by the pandemic compared to students whose parents also attended college or those who began college as freshmen.

The pandemic's impact on student learning was largely driven by the sudden shift to remote instruction. Literature on remote learning shows various approaches including fully remote, software-assisted, and hybrid learning^[see @escueta_education_2017 for a comprehensive review.]. While online learning offers reduced costs in delivering education and wider accessibility, research indicates mixed results. Studies using randomized trials found that students in remote formats generally performed worse than those in traditional settings [@joyce_does_2015, @alpert_randomized_2016]. @bettinger_virtual_2017 and @cacault_distance_2021 found that online learning particularly disadvantaged lower-performing students. Multiple analyses have demonstrated that online courses lead to lower completion rates, grades, and persistence [@jaggars_how_2016; @xu_promises_2019].

Several studies attempt to use the pandemic as an exogenous shock to measure the impact of remote learning on  college students’ outcomes. For instance, in their study, @altindag_is_2021 analyzed administrative data from a public university and employed a fixed effects model. They examine the effect of the change in  learning modality due to the pandemic on students’ learning outcomes. They found that the online instruction mode led to lower grades and an increased likelihood of course withdrawal. Students who have had greater exposure to in-person instruction have a lower likelihood of course repetition, a higher probability of graduating on time, and achieving a higher graduation GPA. Additionally, they observed that the difference in student performance between in-person and online courses tended to diminish over time in the post-pandemic era. 

In the fall of 2020, @kofoed_zooming_2021 randomized 551 West Point students in a required introductory economics course across twelve instructors into either an online or in-person class. They found that final grades for online students dropped by 0.215 standard deviations. This result was apparent in both assignments and exams and was largest for academically at-risk students. Additionally, using a post-course survey, they found that online students struggled to concentrate in class and felt less connected to their instructors and peers. They conclude that the shift to online education had negative effects on learning. Using data on Virginia community college students, @bird_negative_2022 applied a difference-in-differences research design leveraging instructor fixed effects and student fixed effects to estimate the impact of the transition to online learning due to the pandemic. Their results show a modest negative impact of 3% - 6% on course completion. Additionally, their findings suggest that faculty experience in delivering online lectures does not mitigate the negative effects. In their exploratory analyses, they find minimal long-term effects of the switch to online learning. 

A comprehensive study by @bonacini_unraveling_2023, disentangle the channels through which the pandemic affected students. They use admin data from 2018-2021 of 36,000 university students in Italy who took about 400,000 exams during this period. They examine the overall effect of the pandemic on students’ exam scores in different courses. Additionally, they explore the effect of the transition to remote learning by using COVID as an exogenous shock with a difference-in-differences design. Their findings show that during the pandemic, students performed better, with an increase in exam scores. However, the abrupt move to remote learning decreased students’ exam scores.

Studies using survey data on students discussed above have found a negative impact of COVID-related disruptions on academic performance. However, studies that use measured outcomes to evaluate academic performance report mixed results, especially immediately after the pandemic began [@bird_negative_2022; @bonacini_unraveling_2023]. One reason for this might be that many institutions temporarily implemented policies to reduce the burden on students during the pandemic, particularly due to the sudden transition from traditional to fully remote learning. Instructors were likely more lenient in setting exam questions and grading, and more willing to accommodate students than before the pandemic. The sudden move to remote learning could have also created more opportunities for misbehavior by students during exams. For instance, @rodriguez-planas_covid-19_2022, using data from an urban public college in NYC found that lower-income students were 35 percent more likely to utilize the flexible pass/fail grading policy. While no GPA advantage is observed among top-performing lower-income students, in the absence of the flexible grading policy these students would have seen their GPA decrease by 5% relative to their pre-pandemic mean.

The literature has provided valuable insights into the impact of the COVID-19 pandemic on undergraduates. However, several issues remain to be addressed. Many studies rely on self-reported survey data, which may not accurately capture the true extent of learning loss [@aucejo_impact_2020; @rodriguez-planas_hitting_2020]. I identify major limitations in these recent studies. First, using course completion rates, course GPAs, or end-of-semester GPAs to measure academic outcomes immediately after COVID-19 hit in March may not accurately reflect students' actual learning or learning loss. Second, the pandemic-driven sudden transition to new instruction modalities likely changed assessment methods as instructors and students took time to adjust to the situation. The difficulty of exams immediately after the adjustment may not have been the same as pre-COVID exams, contributing to inaccurate measurement of learning loss. Additionally, the implementation of flexible grading policies may have biased the effect of the pandemic on course GPA or course completion rates. I contribute to the literature in two ways. To address these limitations, I analyze students' performance on common exams before and during the pandemic. To eliminate variation due to changes in the difficulty of exams during the pandemic, I examine students' performance on nearly identical questions from exams before and during the pandemic to measure learning loss.

# Data

The data for this study are drawn from two primary sources covering the academic years 2019 through 2022. The first source records student performance on common departmental final examinations for the *Introductory Microeconomics* course at a large public university in New York City. This course is offered every semester, taught by multiple instructors, and at least 700 students enroll annually.

The department offers the course through three modalities. *Hybrid* sections meet twice weekly, comprising one in-person session and one fully remote session. *Online* sections are conducted entirely remotely using software. In 2019 (Spring and Fall), the course was primarily offered in the hybrid mode, with one large online section. Following the onset of the COVID-19 pandemic—spanning Fall 2020 through Spring 2022—offerings were exclusively online or hybrid, with the exception of a single in-person section in 2022. I do not include those students in the analyses to facilitate the comparison between the efficacy of hybrid and online learning modes.

Although the course involves multiple instructors and teaching modalities, assessment is standardized; all enrolled students are required to complete a common, multiple-choice final examination with a maximum score of 40 points. Leveraging performance on common exams eliminates potential bias arising from heterogeneity in instructor-specific testing difficulty. The dataset includes answer sheets for all students who attempted these exams, providing the final score, item-level performance, instructor identifiers, and the course learning mode. Data for the Spring 2020 semester were unavailable.

The first measure is the aggregate score on the common final examination, converted from a raw maximum of 40 points to a standard 0–100 scale. This standardized measure provides a more consistent signal of learning than course GPA or withdrawal rates, which were potentially confounded by flexible grading policies adopted during the pandemic.

To exploit the granularity of the data, I also construct a second measure based on item-level performance. Because the final exams are departmental, I am able to match identical or nearly identical questions appearing in exams administered both pre- and post-pandemic onset. While the department utilizes two versions of the exam to deter cheating (differing only in question order), the content remains constant. I manually identified and matched 35 unique pairs of questions across the pre-pandemic and pandemic periods. This allows for the construction of a binary outcome variable, *correct*, which takes a value of 1 if a student answered the matched question correctly and 0 otherwise.

The second data source consists of institutional administrative records for all students enrolled in *Introductory Microeconomics* during the relevant semesters. This dataset includes a rich set of covariates, including gender, race, age, transfer status, enrollment intensity (part-time vs. full-time), native language, and class standing (freshman through senior).

By merging these administrative records with the examination data, I construct a comprehensive dataset linking student characteristics to standardized performance metrics. This merger also incorporates exam-level characteristics, such as learning modality, course instructor, exam version, and the semester of administration. To my knowledge, this is the first dataset to facilitate a granular examination of COVID-19's impact on student performance using standardized, item-level outcomes.

The final analytical sample comprises 4,655 unique students enrolled in the course. For the granular analysis using matched exam questions, the dataset expands to 47,589 student-question observations. Each observation represents a student-question pair indicating whether the specific item was answered correctly. For the majority of the sample, I utilize the cumulative GPA recorded prior to the start of the semester. If the pre-semester GPA is unavailable, I substitute it with the GPA calculated at the end of the concurrent semester. In cases where both values are missing, I impute the missing value using the mean GPA of the student cohort for that specific semester.

To estimate the impact of the pandemic on student learning, I employ a series of Ordinary Least Squares  regressions. I first examine aggregate performance on the common final examination, followed by an analysis of item-level performance using matched questions across pre-pandemic and pandemic semesters. My primary empirical strategy estimates the effect of the pandemic on student outcomes using the following specification:

\begin{equation}
y_{i,c,t} = \delta P_{t} + \beta X_{i,c,t} + \gamma_{c} + \alpha_{s} + \epsilon_{i,c,t}
\end{equation}

$y_{i,c,t}$ represents the academic outcome for student $i$ in class taught by instructor $c$ during semester $t$. I analyze two distinct outcomes for $y$. In the first set of regressions, $y$ is the students' aggregate score on the common final examination, scaled to a 0–100 range. In the second set of regressions,  I employ a linear probability model where $y$ is a binary indicator equal to 1 if the student answered a specific matched question correctly, and 0 otherwise. For this item-level analysis, the subscript $q$ is added to denote the specific question pair, such that the outcome is denoted as $y_{i,c,q,t}$.

The variable of interest is $P_{t}$, a binary indicator for the pandemic period, equal to 1 for any semester after Fall 2019 (i.e., Spring 2020 onward) and 0 otherwise. The coefficient $\delta$ captures the average effect of the pandemic on student performance.

$X_{i,c,t}$ includes student-level controls to account for demographic and academic heterogeneity. These include dummy variables for race (Black, Asian, non-White Hispanic, and Other, with White as the reference group), gender (equal to 1 if female), and class standing (equal to 1 if the student is a freshman or sophomore). To control for baseline ability, I include the student's cumulative GPA prior to the start of the course. The model also includes instructor fixed effects ($\gamma_{c}$) and session fixed effects ($\alpha_{s}$) to control for time-invariant instructor characteristics and semester-specific shocks unrelated to the pandemic. Standard errors are robust to heteroskedasticity.

## Identification of Differential Impact of COVID-19 on Low vs High GPA Students

I further examine whether the pandemic differentially affected students based on their prior academic performance. I classify students into "Low GPA" and "High GPA" groups based on the sample median cumulative GPA of 3.32. Students below this threshold are defined as low-GPA, while those at or above are defined as high-GPA. I estimate the following interaction model:

\begin{equation}
y_{i,c,t} = \delta P_{t} + \phi L_{i} + \mu P_{t} \times L_{i} + \beta X_{i,c,t} + \gamma_{c} + \alpha_{s} + \epsilon_{i,c,t}
\end{equation}

Here, $L_{i}$ is a binary indicator equal to 1 if student $i$ is in the low-GPA group. The coefficient of interest is $\mu$, which captures the differential impact of the pandemic on low-performing students relative to their high-performing peers. A negative $\mu$ would indicate that the pandemic exacerbated inequality in learning outcomes. This specification mirrors the baseline model but excludes continuous cumulative GPA from the vector $X_{i,c,t}$, as it is captured by the group classification.

## Identification of the Effect of Sudden Transition to Remote Learning

Finally, I isolate the effect of the sudden, forced transition to remote learning. While the aggregate pandemic effect ($\delta$ in Equation 1) captures broadly defined disruptions, a key mechanism was the shift in instructional modality. Prior to the pandemic, the department offered the course in two distinct modes: Hybrid (one in-person and one online session weekly) and Online (fully remote). The onset of the pandemic in March 2020 forced all hybrid sections to transition abruptly to a fully remote format. I exploit this variation using a difference-in-differences framework to estimate the impact of transitioning from hybrid to remote learning, relative to students who were already enrolled in fully online sections:

\begin{equation}
y_{i,c,t} = \delta P_{t} + \phi O_{i} + \mu P_{t} \times O_{i} + \beta X_{i,c,t} + \gamma_{c} + \alpha_{s} + \epsilon_{i,c,t}
\end{equation}

In this specification, $O_{i}$ is an indicator equal to 1 if the student originally enrolled in an online section and 0 if they enrolled in a hybrid section. The interaction term $\mu$ tests whether students who selected into online learning—and thus experienced less disruption in modality—performed differently during the pandemic compared to those forced to switch from hybrid to remote instruction. Control variables remain consistent with the baseline specification, excluding the instruction mode indicator which is subsumed by the difference-in-differences terms.

# Results

```{r data-load, echo = FALSE, warning=FALSE, message=FALSE, include = FALSE}
### Loading the question level data ------------------------------------------

question_url <- "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/question_level.rds"
question_level <- read_rds(question_url) |>
  mutate(on = online) ## This is to ensure there are no glitches in the code where two panels of results are merged. A bad fix, but works!

### Loading the exam level data ------------------------------------------
## These scores are out of a possible 40 points.

exam_url <- "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/exam_level.rds"
exam_level <- read_rds(exam_url) |>
  mutate(on = online) ## This is to ensure there are no glitches in the code where two panels of results are merged. A bad fix, but works!


### Loading the course withdrawal data ------------------------------------------
## It has share of students who took withdrawal, CR, NC options.
full_grades <- read_dta(
  "https://raw.githubusercontent.com/amanojas/COVID-19-and-Academic-Performance/main/data/grades-by-sem.dta"
) # s2020 available
full_g <- full_grades |>
  mutate(
    year = if_else(
      semester == "S19" | semester == "F19",
      2019,
      if_else(
        semester == "S20" | semester == "F20",
        2020,
        if_else(semester == "S21" | semester == "F21", 2021, 2022)
      )
    )
  ) |>
  mutate(
    session = factor(if_else(
      semester %in% c("S19", "S20", "S21", "S22"),
      "S",
      "F"
    ))
  ) |>
  mutate(session = fct_relevel(session, c("S", "F")))

```

## Average Course GPA Across Semesters in ECO 1001

```{r gpa-across-semesters-eco-1001, echo = FALSE, warning=FALSE, message=FALSE, fig.cap= "(#fig:gpa-across-semesters-eco-1001) Average Course GPA in ECO 1001 across semesters", fig.pos="H",  fig.align = "center"}
# Define semester labels
gpa_labels <- c(
  "Spring 2019",
  "Fall 2019",
  "Spring 2020",
  "Fall 2020",
  "Spring 2021",
  "Fall 2021",
  "Spring 2022"
)

# Summarize GPA statistics by semester
gpa_summary <- full_g |>
  dplyr::select(year, session, coursegpa) |>
  dplyr::group_by(year, session) |>
  dplyr::summarise(
    avg_gpa = mean(coursegpa, na.rm = TRUE),
    sd_gpa = sd(coursegpa, na.rm = TRUE),
    n = dplyr::n(),
    .groups = 'drop'
  ) |>
  dplyr::mutate(
    margin_error = 1.96 * (sd_gpa / sqrt(n)),
    semester = factor(seq_along(year), labels = gpa_labels)
  )

# Plot the results
ggplot(gpa_summary, aes(x = semester, y = avg_gpa, label = round(avg_gpa, 2))) +
  geom_bar(stat = "identity", width = 0.3, fill = "#009E73", alpha = 0.8) +
  geom_errorbar(
    aes(ymin = avg_gpa - margin_error, ymax = avg_gpa + margin_error),
    width = 0.1,
    color = "darkblue"
  ) +
  geom_text(
    size = 3,
    vjust = -2.5,
    color = "black"
  ) +
  ylim(0, 4) +
  labs(
    x = NULL,
    y = "Course GPA",
    title = ""
  ) +
  scale_x_discrete(
    labels = gpa_labels,
    guide = ggplot2::guide_axis(n.dodge = 2)
  ) +
  theme_minimal(base_size = 10)
```

A central premise of this study is that conventional metrics commonly cited in the literature---such as course completion rates, withdrawal rates, and Grade Point Averages (GPA)---may fail to accurately capture academic performance during the pandemic period. The abrupt transition in instructional modality likely induced behavioral responses from instructors, including increased leniency in grading standards to accommodate the disruptive environment. Furthermore, the shift to remote assessment may have introduced opportunities for academic misconduct, thereby inflating scores. Consequently, relying solely on GPA may mask the true negative impact of the pandemic on human capital accumulation, as these institutional policy shifts and grading accommodations can overshadow actual declines in content mastery.

Figure 1 presents the temporal evolution of unadjusted average GPAs in the introductory microeconomics course. The data reveal a sharp discontinuity in Spring 2020, coinciding with the transition to online instruction. This increase in average grades contradicts contemporaneous survey evidence suggesting that students faced substantial personal and academic hardships during this period [@aucejo_impact_2020; @rodriguez-planas_hitting_2020]. If GPAs were a consistent proxy for learning, one would expect a decline in grades reflecting these documented struggles. Instead, while GPAs moderated slightly in Fall 2020 and Spring 2021, they remained persistently elevated above pre-pandemic levels until after Fall 2021, suggesting a structural break in grading norms rather than an improvement in student performance.

## Withdrawal Rate Across Semesters in ECO 1001

Another possible mechanism leading to the opposing change in measured performance is that the institution in consideration, like many other academic institutions, adopted a flexible grading policy to help students face the challenges due to the pandemic. This policy aimed to reduce the burden on students by providing three options up until the last day of the semester. The first option, Credit (CR), allowed students to pass the course with credit, though their grade wouldn't affect their GPA. The second, No Credit (NC), let students complete the course without credit, allowing them to retake it later without any record of their withdrawal. The third was the standard course withdrawal option.

Figure 2 shows the unadjusted withdrawal rates across semesters in ECO 1001. The course withdrawal rate decreased to 0.83% in fall 2019, down from 4.97% in spring 2019. However, it increased again to 3.92% in spring 2020, a semester heavily influenced by the onset of the pandemic. Despite the pandemic, the withdrawal rate was kept relatively low due to the introduction of a flexible grading policy by the college. As shown in the figure, 29.75% of students enrolled in ECO 1001 chose the CR option, while 5.53% chose NC. Because of this flexibility, only 3.92% of students opted for a standard withdrawal in spring 2020. The withdrawal rate increased to 6.65% in fall 2020 and remained roughly at that level, reaching around 8% in spring 2022. It is worth noting that the flexible grading policy was not implemented after spring 2020. Using course GPA or course completion rate in presence of a flexible grading policy may not give me a clear effect of the pandemic on students’ academic outcomes and their learning loss.

## Descriptive Statistics

Table 3.1 outlines the sample characteristics before and after the pandemic. The sample includes 4,598 students enrolled in the course. The pre-Covid period covers observations from spring and fall 2019. The post-Covid data includes students enrolled in fall 2020, spring 2021, fall 2021, and spring 2022. The table reports the pre-Covid and post-Covid averages of the variables as well as differences in their means.

On average, unadjusted difference in exam scores of the students in the common final exams is 0.562 points. This difference is not statistically significant. In case of performance on nearly identical questions, the average probability of answering the question, unadjusted, is about 7.6 percentage points less in post-Covid exams relative to pre-Covid exams. The difference is statistically significant at 1% level. Regarding student demographics, there has been an increase in the proportion of Hispanic students in the course from 13.3 percent before the pandemic to 18.9 percent after. The enrollment proportion for Asian students has decreased, with a difference of -5.5 percent. The proportion of Black students has remained roughly the same before and after the pandemic, with the small difference not being statistically significant. The difference in enrollment for students of all races except for Black are statistically significant at the 1 percent level. Before the pandemic, around 35 percent of the students were enrolled in fully online classes. However, in the post-Covid period, about 59 percent of students chose fully online classes over hybrid classes. Notably, all students enrolled in this course took fully remote classes during the fall 2020 and spring 2021 sessions. In contrast, during fall 2021, all students were enrolled in hybrid classes for the course. By spring 2022, both hybrid and online classes were available.

In the post-pandemic period, students are nearly a year younger than in the pre-pandemic period, a difference that is statistically significant at the 1 percent level. The proportion of part-time students has decreased since 2019. The proportion of students whose native language is not English has also decreased significantly from 58.1 percent to 43.1 percent. Most students taking the introductory microeconomics course are freshmen or sophomores. Their proportion has increased by 10 percentage points in the post-pandemic period compared to the pre-pandemic period. A crucial control variable in this study is the students' GPA, for which I use their cumulative GPA from before the semester in which they enrolled in the course started. Some observations have missing values. If a student's cumulative GPA at the start of the semester is missing, I replace it with their GPA at the end of the semester. If a student's cumulative GPA before or after the semester is missing, I impute the value using the mean GPA of the semester in which the student enrolled in the course for further analyses.


## Baseline Specification

Table 3.2 presents the results of the baseline specification. As stated earlier, student performance was measured using two outcome variables. The coefficients with standard errors are reported. Also reported below the standard errors are standardized coefficients in brackets. In the first two columns, the outcome variable is the student’s exam score on the common final exam. It appears from a simple model in the first column that performance measured using the exam score, decreased in the post-pandemic period. Looking at the first column, on average, in the post-pandemic period, the exam score decreased by a point (or 0.02 standard deviations), although the coefficient is not statistically significant. Column 2 shows the results by semester using dummy variables, with spring and fall 2019 combined as the benchmark category. Due to limited pre-pandemic observations, I combined spring and fall 2019 data into a single period. When the pandemic struck, the score increased by 1.35 points in fall 2020 compared to exam scores in 2019, but the coefficient is not statistically significant. The scores decreased sharply in spring 2021 by 5.75 points or 0.37 standard deviations below the mean score. Mean scores increased in fall 2021 before decreasing in spring 2022 by 6.7 points (or 0.43 standard deviations). 

Columns 3-4 present the results from linear probability models, where the outcome variable is binary since I look at the students’ performance on matched questions from pre and post pandemic final exam.  For a full period post pandemic, the probability of students answering a similar question from pre pandemic exam decreases by 1.5 percentage points. Analyzing the results across semesters, immediately after the pandemic struck, I see a sharp decrease in the probability of students answering a nearly identical question correctly in fall 2020 compared to the common final exams in 2019. The probability of answering the nearly identical question in fall 2020 decreased by 10 percentage points or 0.21 standard deviations below the mean probability compared to that of in 2019. The performance appeared to improve in subsequent semesters, with the probability of answering the nearly identical question from pre-pandemic common exams during  the pandemic decreased by about 8 percentage points in spring 2022. In all regressions, I control for students' demographic characteristics, including race and gender, as well as other factors such cumulative GPA and their part-time student status.  In addition to that, in all regressions, I control for the gpamiss variable to see if the results change due to mean imputation of missing GPA values. The results do not appear to change due to that.

I also examine closely the differential effect of the pandemic based on students' GPA quartiles for both exam scores and matched questions data. For the exam scores dataset, the GPA quartiles are constructed as follows: first quartile: GPA ≤ 3.01, second quartile: 3.01 < GPA ≤ 3.37, third quartile: 3.37 < GPA ≤ 3.71, and fourth quartile: GPA > 3.71. For matched questions data, the GPA quartiles are constructed as follows: first quartile: GPA ≤ 3.08, second quartile: 3.08 < GPA ≤ 3.32, third quartile: 3.32 < GPA ≤ 3.68, and fourth quartile: GPA > 3.68. Using students in the fourth GPA quartile as a benchmark group, I can examine how the pandemic affected students in other quartiles.




## Impact of COVID on Low GPA Students 

Panel A in table 3.4 show the results of differential impact of the pandemic on the performance of students with low GPA compared to their high GPA counterparts. As explained earlier, I define low GPA students with GPA less than median GPA of 3.2. Columns 1 and 2 show results from OLS regressions with final exam scores as the outcome variable. On average, low GPA students score 11.4 points lower than high GPA students on the common final exam. In column 2, I include an interaction term that combines the low GPA dummy with a dummy for the post-COVID period. This is similar to a standard difference-in-difference estimate of the pandemic's effect on the performance of low GPA students relative to high GPA students, where I assume the pandemic did not affect high GPA students’ performance. I see that due to the pandemic, the average exam scores of low GPA students decreased by 3.3 points (or 0.04 standard deviation) relative to high GPA students.

Comparing these results to those from linear probability models in columns 3-4, I see a statistically significant reduction in the performance of low GPA students. This is measured by their ability to answer nearly identical questions in exams post-pandemic from the pre-pandemic common exams. In column 3, I see that, on average, low GPA students are 13.2 percentage points (or 0.13 standard deviations) less likely to answer a similar question compared to their high GPA counterparts. In column 4, the coefficient on an added interaction term suggests that post-pandemic, low GPA students are 3.3 percentage points less likely to answer a similar question from pre-pandemic common exams compared to high GPA students. The coefficient is statistically significant at the 1% level.

## Abrupt Transition to Remote Learning

Panel B in table 3.4 displays the results of the impact of the pandemic-induced abrupt transition to remote learning from the pre-pandemic hybrid mode of learning. Columns 1-2 present the results of OLS models where the outcome variable is the final exam scores of the students. On average, students enrolled in online classes score about 1.9 points less than those in hybrid classes, controlling for the COVID period. In column 2, I interact a dummy variable for the COVID period with a dummy variable for remote learning. The coefficient on the interaction term could be understood as a difference in differences estimate of suddenly transitioning to online classes from hybrid classes. For the *introductory microeconomics* course, pre-COVID, the department offered both hybrid and online classes. When the pandemic hit, the department followed the nationwide policy of abruptly transitioning to online classes. The coefficient on the interaction term thus presents the impact of this sudden shift to online learning from hybrid learning on students' performance. The estimate is -5.432  (or -0.06 standard deviations) and is statistically significant at the 1% level. 

Columns 3-4 show the results of linear probability models with binary outcome variable which is 1 if a student answers the question correctly and 0 otherwise.  Column 3 shows that on average, accounting for dummy variable for the pandemic, students enrolled in online course are 7.3 percentage points less likely to answer a nearly identical question from common exams from pre pandemic period in post pandemic exams. Column 4 is a classic difference-in-differences specification. Surprisingly, the impact of a sudden transition from hybrid to online learning increased the students’ probability of answering a similar question from pre-pandemic common exams in the post-pandemic period by 5.6 percentage points. 

In table 3.4, in panel A, all regressions include the following control variables: instruction mode, gender, race, and part-time status of the student. In panel B, all regressions include the following control variables: cumulative GPA, gender, race, and part-time status of the student. All regressions also include a dummy variable, gpamiss, which is 1 if cumulative GPA is imputed using the mean and 0 otherwise. All regressions include session fixed-effects and course instructor fixed-effects.



## Dynamic Effects

I am also interested in examining the differential impact of the pandemic on the outcomes of high and low GPA students across the semesters. I interact the low GPA with separate time dummies for all semesters, with spring 2019 and fall 2019 combined as the benchmark category. This allows me to explore how the outcome differences between low GPA and high GPA students evolve over time. I also perform the same exercise to explore the impact of abrupt transition to online mode of learning across the semesters. I interact a dummy variable for the online mode of learning with all semester dummies, with the same benchmark category.

In figure 3.3, the outcome variable is scores on the common final exam. The top-left panel illustrates the long-term impact of COVID-19 on exam scores of low-GPA students compared to high-GPA students. There's a sharp decline in exam scores for low-GPA students in Fall 2020. Although their performance improves over time, a gap persists. The top-right panel illustrates the impact of the sudden transition to online learning on exam scores across different semesters. Immediately after the COVID-19 hit, transition to online classes decreased exam scores but recovered after one semester suggesting gradual adaptation to new learning environment.

A similar pattern emerges with matched question data used to measure students' academic outcomes. The mean probability of answering a nearly identical question post-pandemic compared to pre-pandemic exam decreases sharply for low-GPA students immediately after COVID-19 hit (bottom-left panel). It did not appear to recover by spring 2022. The average probability of answering a similar question correctly decreases due to the sudden transition to online classes but then increases to the levels seen before the pandemic (bottom-right panel). 

## Effects Due to Heterogenity in Difficulty of Questions

In this section, I examine the pandemic's effect on students' performance by analyzing the mean probability of answering nearly identical questions before and during the pandemic, taking into account the question difficulty. As previously mentioned, all students enrolled in ECO 1001 take a common final exam. Course instructors categorized questions as either *easy* or *hard* based on their difficulty level.  Since, I am matching question from pre pandemic exams to exams during the pandemic, I classify questions as *hard* if they were labelled *hard* by the instructors from the pre-pandemic exams, and the follow the same logic for the easy questions. 

Results from panel A in table 3.5 show that on average low GPA students are about 15 percentage points (0.15 standard deviations) less likely to answer a hard question correctly compared to high GPA students. Due to the pandemic, low-GPA students' mean probability of answering nearly identical hard questions correctly decreased by 4.5 percentage points (or 0.02 standard deviations) compared to high-GPA students. For easy questions, their performance decreased by 1.6 percentage points, though this estimate is not statistically significant. In Panel B, I do a similar analysis for students enrolled in online relative to hybrid classes.  On average, students enrolled in online classes are 7.2 percentage points (or 0.07 standard deviations) less likely to answer a hard question correctly compared to students enrolled in hybrid classes. Following the abrupt transition from pre-pandemic hybrid learning to online mode, mean probability of answering nearly identical hard questions correctly increased by just over 12 percentage points (0.06 standard deviations). For easy questions, the estimate is not statistically significant.

# Conclusion

In this essay, I examine the pandemic's influence on the academic performance of students by analyzing their results in the common exams for introductory microeconomics course at a large public university in New York City.  I advance the literature by providing estimates of learning loss in college students due to pandemic that are more reliable than current estimates. I use two outcome measures to evaluate students' academic performance and argue that these outcome choices are more appropriate than the existing outcome measures such as course completion rate, course GPA, or semester GPA used in the literature on the impact of COVID on students' academic performance. First, I analyze students' scores on common final exams administered at the institution from 2019 to 2022, excluding spring 2020 due to lack of data availability for that semester. Acknowledging the fact that difficulty of exams may have changed during the pandemic, I use 35 pairs of questions matched  from these common final exams to measure changes in the students’ average probability of answering nearly identical questions from the exams conducted before and during the pandemic to eliminate the variation from exam difficulty. I find an overall negative impact of the pandemic on students’ outcomes. Students’ scores went down by a point (or 0.02 standard deviations) in the full pandemic period (2020-2022), although the coefficient is not statistically significant. Students’ average probability of answering similar questions from the common exams before the pandemic went down during the pandemic by 1.5 percentage points. This clear evidence of learning loss, I argue, is not affected by the flexible grading policy. The extent of learning loss was greater in fall 2020 and gradually lessened through fall 2021, after which it stabilized. 

I also examine the differential impact of the pandemic on the outcomes of students with low GPA compared to those with high GPA. My findings suggest that on average low GPA students have a 3.3 percentage point lower average probability of correctly answering similar questions compared to high GPA students during the pandemic. This accounts for a broad range of student characteristics and incorporates instructor and session fixed effects, indicating a significant differential impact on low GPA students. While using students' scores from common exams as the outcome variable, I find that low GPA students on average scored 3.23 points (or 0.04 standard deviations) less in the common exams compared to high GPA students during the pandemic. In the long term, although this difference decreases, it does not return to the pre-pandemic level by spring 2022. Additionally, I examined the pandemic's effects across GPA quartiles and found that students in the lowest quartile of GPA distribution were 4.1 percentage points (0.02 standard deviations) less likely to correctly answer nearly identical questions from pre-pandemic exams during the pandemic. This analysis supports the hypothesis that low GPA students, on average, suffered greater learning loss due to the pandemic compared to high GPA students.

Furthermore, I explore an important channel: the sudden shift to online classes, through which the pandemic affected students' academic outcomes. I find that abruptly moving to online classes due to the pandemic reduced students’ final exam scores by 5.43 points. In case of matched questions data, the mean probability of answering a similar question before and after suddenly moving to online classes increased by 5.6 percentage points. Interacting the semester dummies with a dummy for online variable, I find that the abrupt transition to online classes reduced the average probability of answering a similar question correctly before and during pandemic before returning to pre-pandemic levels. The same pattern is observed in case of exam scores as outcome variable. To examine how sensitive these estimates of learning loss are to question difficulty in the matched questions data, I provide results from separate analyses using easy as well as hard questions. During the pandemic, low-GPA students' mean probability of answering nearly identical hard questions decreased by 4.5 percentage points relative to their high-GPA counterparts. I found no statistically significant effect for easy questions.  When examining the effect of abrupt transition to remote classes, I found that students scored just over 12 percentage points higher on hard questions after moving online, while showing no statistically significant difference on easy questions.

Overall, I find negative effects of the pandemic on students' academic performance that align directionally with the current literature. My unique matched questions data allows me to eliminate bias in the estimates that arose from flexible grading policies implemented immediately after the pandemic hit educational institutions nationwide. I do, however, acknowledge that my estimates may not account fully for potential cheating by students, especially in the initial months following a sudden transition to remote classes. The implications of learning loss due to the pandemic could be significant. On one hand, students' GPAs, both course-specific and overall, did not change much or even increased in some cases during the pandemic, giving the impression of better performance. On the other hand, evidence from student surveys shows that students faced hardships and challenges in learning during this time. In my study I provide evidence of learning loss which is consistent with students’ negative experiences during the pandemic. In future, any decision to suddenly switch to remote learning during a complex situation should be carefully considered before implementation.

\pagebreak

# References

::: {#refs}
:::

\pagebreak

# Appendix

